---
title: "PRECIS-2 Inter-Rater Reliability Analysis"
author: "Aarian Bhakoo"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: yes
    number_sections: yes
---

<!-- Setup and configuration -->
```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.align = "center"
)

# Required packages
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(knitr)
library(krippendorffsalpha)

# Display numerical results consistently
options(
  digits = 4,
  scipen = 999
)

set.seed(2026)
```

This report reproduces the prespecified inter-rater reliability analyses for
the finalized set of 237 paired PRECIS-2 ratings. Codes 1--5 represent the
ordinal pragmatism scale; code 6 represents Missing and code 7 represents Not
Applicable. The analysis implements an alpha estimation approach reviewed by
B.K. and applies it to the final dataset and agreement definitions.


# Data Setup and Validation
<!-- This step relates numerical codes 1-7 to nominal and ordinal classifications
on the Modified PRECIS-2 scale, and defines the maximum level of disagreement
for disagreement weighting later in the analysis-->
```{r defining-agreement}
# Permitted rating codes:
#   1-5 = ordinal PRECIS-2 scores
#   6   = Missing
#   7   = Non-Applicable

VALID_CODES <- 1:7
ORDINAL_CODES <- 1:5
MISSING_CODE <- 6
NON_APPLICABLE_CODE <- 7
NOMINAL_CODES <- c(MISSING_CODE, NON_APPLICABLE_CODE)

# Maximum ordinal disagreement is defined using scores 1 and 5.
MAX_ORDINAL_DIFFERENCE <- 5 - 1
MAX_DISAGREEMENT <- 1
```


```{r loading-files}
# Locate the repository whether the report is rendered from the project root
# or directly from the analysis directory in RStudio.
project_root <- if (dir.exists(file.path("data", "derived"))) {
  normalizePath(".", mustWork = TRUE)
} else if (dir.exists(file.path("..", "data", "derived"))) {
  normalizePath("..", mustWork = TRUE)
} else {
  stop("Could not locate the repository's data/derived directory.")
}

repo_path <- function(...) file.path(project_root, ...)

# Read the final dataset ("precis2_irr_final.csv")
# The column aliases below preserve the statistician-reviewed analysis object names.
final_data_file <- repo_path(
  "data", "derived", "precis2_irr_final.csv"
)

final_data <- read.csv(
  final_data_file,
  check.names = FALSE,
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
) |>
  arrange(observation_id)

# Creating specific views of the independent and consensus ratings
ratings_tidy <- final_data |>
  transmute(
    observation_id, trial_order, trial, domain_number, domain,
    observer_1 = aarian_score,
    observer_2 = merrick_score
  )

consensus_tidy <- final_data |>
  transmute(
    observation_id, trial, domain_number, domain,
    consensus = consensus_score
  )

rater_names <- c("A.B.", "M.Z.")
censored_rater_names <- rater_names

# Construct units-by-raters matrix and transposed raters-by-units form
# Both orientations are retained
ratings_units_by_raters <- ratings_tidy |>
  select(observer_1, observer_2) |>
  mutate(across(everything(), as.numeric)) |>
  as.matrix()

colnames(ratings_units_by_raters) <- rater_names
rownames(ratings_units_by_raters) <- ratings_tidy$observation_id
ratings_matrix <- t(ratings_units_by_raters)

# For "Censored", treat Missing (6) and Non-Applicable (7) as
# absent ratings, then create the corresponding matrix
censored_units_by_raters <- ratings_units_by_raters
censored_units_by_raters[
  censored_units_by_raters %in% NOMINAL_CODES
] <- NA_real_
censored_matrix <- t(censored_units_by_raters)
```


<!-- This step ensures that the data were loaded correctly, without
unexplained transformations or losses -->
```{r validation}
# Confirm that the main dataset contains only valid codes or genuine NA values.
required_rating_columns <- c(
  "observation_id", "trial_order", "trial", "domain_number", "domain",
  "observer_1", "observer_2"
)
required_consensus_columns <- c(
  "observation_id", "trial", "domain_number", "domain",
  "consensus"
)

# Error handling
if (!all(required_rating_columns %in% names(ratings_tidy))) {
  stop("The finalized ratings table is missing one or more required columns.")
}

if (!all(required_consensus_columns %in% names(consensus_tidy))) {
  stop("The finalized consensus table is missing one or more required columns.")
}

if (nrow(ratings_tidy) != 237L || nrow(consensus_tidy) != 237L) {
  stop("Both finalized tables must contain exactly 237 observations.")
}

if (anyDuplicated(ratings_tidy$observation_id) ||
    anyDuplicated(consensus_tidy$observation_id)) {
  stop("Observation identifiers must be unique within each finalized table.")
}

if (!identical(ratings_tidy$observation_id, consensus_tidy$observation_id)) {
  stop("Ratings and consensus observations are not aligned.")
}

if (!identical(ratings_tidy$trial, consensus_tidy$trial) ||
    !identical(ratings_tidy$domain_number, consensus_tidy$domain_number)) {
  stop("Trial or domain labels differ between ratings and consensus tables.")
}

if (!identical(sort(unique(ratings_tidy$domain_number)), 1:10)) {
  stop("The analysis must contain the ten final PRECIS-2 domains.")
}

if (length(unique(ratings_tidy$trial)) != 24L) {
  stop("The finalized analysis must contain 24 trials.")
}

# Confirmed that prespecified exclusions detailed in
# "precis2-irr/data/exclusions.csv" are absent from the final dataset
excluded_rows_present <- ratings_tidy |>
  filter(trial == "Thompson (2000B)", domain_number %in% 7:9)
# handle errors
if (nrow(excluded_rows_present) > 0L) {
  stop("Excluded Thompson (2000B) Domains 7-9 are present in the analysis.")
}

# Stop if any rating falls outside 1-7 (e.g., 8)
observed_codes <- sort(unique(as.vector(ratings_matrix)))
observed_codes <- observed_codes[!is.na(observed_codes)]

invalid_codes <- setdiff(observed_codes, VALID_CODES)

if (length(invalid_codes) > 0) {
  stop(
    "The main ratings file contains invalid codes: ",
    paste(invalid_codes, collapse = ", ")
  )
}

# Confirm that the censored dataset contains only scores 1-5 or NA.
observed_censored_codes <- sort(unique(as.vector(censored_matrix)))
observed_censored_codes <-
  observed_censored_codes[!is.na(observed_censored_codes)]

invalid_censored_codes <-
  setdiff(observed_censored_codes, ORDINAL_CODES)

if (length(invalid_censored_codes) > 0) {
  stop(
    "The censored file contains values outside 1-5: ",
    paste(invalid_censored_codes, collapse = ", ")
  )
}

# Confirm that both files contain the same raters and observations.
if (!identical(rater_names, censored_rater_names)) {
  stop("The rater names or rater order differ between the two files.")
}

if (!identical(dim(ratings_matrix), dim(censored_matrix))) {
  stop("The two datasets do not have the same dimensions.")
}

if (nrow(ratings_matrix) < 2) {
  stop("At least two raters are required.")
}

if (!identical(dim(ratings_matrix), c(2L, 237L))) {
  stop("The finalized rater matrix must have 2 raters and 237 observations.")
}

if (sum(apply(ratings_units_by_raters, 1, function(x) {
  any(x %in% NOMINAL_CODES)
})) != 26L) {
  stop("Expected 26 observations containing a previously nominal score.")
}

if (sum(stats::complete.cases(censored_units_by_raters)) != 211L) {
  stop("The censored analysis must contain 211 eligible observations.")
}

# Output report should read
#   Number of Raters: 2
#   Number of Rated Observations: 237
#   Observed Codes: 1, 2, 3, 4, 5, 6, 7
cat("Setup completed successfully.\n")
cat("Number of raters:", nrow(ratings_matrix), "\n")
cat("Number of rated observations:", ncol(ratings_matrix), "\n")
cat(
  "Observed codes:",
  paste(observed_codes, collapse = ", "),
  "\n"
)
```


# Primary Analysis
## Crude Agreement
<!-- Primary Outcome (Global Agreement) Measure 1/2 -->
```{r}
# Calculates crude agreement as:
# (number of agreeing observations) / (number of observations)
# agreeing observations = EXACT match (no partial disagreement)

# Helper Function
crude_agreement <- function(agreement_vector){

  n_total <- length(agreement_vector)
  n_agree <- sum(agreement_vector)

  data.frame(
    Agreements = n_agree,
    Total = n_total,
    Crude_Agreement = n_agree / n_total
  )

}


# [1] Total Agreement
# Agreement only when both raters assigned exactly the same code.
r1 <- ratings_matrix[1, ]
r2 <- ratings_matrix[2, ]
agree_total <- r1 == r2
total_results <- crude_agreement(agree_total)


# [2] Range Agreement
# Scores within one ordinal category are considered agreement.
#
# Exception:
#   5 does NOT agree with 6
#   6 does NOT agree with 7
# 
agree_range <- logical(length(r1))
for(i in seq_along(r1)){
  a <- r1[i]
  b <- r2[i]
  if(a == b){
    agree_range[i] <- TRUE
  } else if(
      abs(a-b) == 1 &&
      !((a==5 & b==6) |
        (a==6 & b==5) |
        (a==6 & b==7) |
        (a==7 & b==6))
  ){
    agree_range[i] <- TRUE
  } else{
    agree_range[i] <- FALSE
  }
}
range_results <- crude_agreement(agree_range)


# [3] Censored Agreement
# Remove observations where either rater recorded
# Missing (6) or Non-Applicable (7).
# Apply Total Agreement definition after parsing
keep <- !(r1 %in% NOMINAL_CODES | r2 %in% NOMINAL_CODES)
r1_censored <- r1[keep]
r2_censored <- r2[keep]
agree_censored <- r1_censored == r2_censored
censored_results <- crude_agreement(agree_censored)


# [4] Bucketed Agreement
# Collapse categories:
#
# 1 & 2 -> 1
# 3     -> 2
# 4 & 5 -> 3
# 6     -> 4
# 7     -> 5
bucket_scores <- function(x){
  case_when(
    x %in% c(1,2) ~ 1,
    x == 3 ~ 2,
    x %in% c(4,5) ~ 3,
    x == 6 ~ 4,
    x == 7 ~ 5
  )

}
bucket1 <- bucket_scores(r1)
bucket2 <- bucket_scores(r2)
agree_bucket <- bucket1 == bucket2
bucket_results <- crude_agreement(agree_bucket)


# Summary Table
agreement_summary <-
  bind_rows(
    Total = total_results,
    Range = range_results,
    Censored = censored_results,
    Buckets = bucket_results,
    .id = "Definition"
  )

# Regression check against the independently verified Step 1 results.
# 1 - Record the accepted crude agreement numerators & denominators
expected_agreement <- tibble(
  Definition = c("Total", "Range", "Censored", "Buckets"),
  Agreements = c(115L, 165L, 110L, 152L),
  Total = c(237L, 237L, 211L, 237L)
)

# 2 - Stop if there have been changes to the finalized crude-agreement count
if (!identical(agreement_summary$Definition, expected_agreement$Definition) ||
    !identical(as.integer(agreement_summary$Agreements),
               expected_agreement$Agreements) ||
    !identical(as.integer(agreement_summary$Total), expected_agreement$Total)) {
  stop("Crude agreement results do not match the finalized Step 1 results.")
}

kable(
  agreement_summary,
  digits = 4,
  caption = "Summary of crude agreement under each agreement definition.",
  booktabs = TRUE
)
```



## Krippendorff's Alpha
<!-- Primary Outcome (Global IRR) Measure 2/2 -->
<!-- Helper Function -->
```{r disagreement-function} 
# Builds a disagreement function for Krippendorff's alpha.
#
# ordinal_positions:
#   Named numeric vector assigning each ordinal category a position.
#
# tolerance:
#   Number of ordinal steps treated as complete agreement.
#   Total/Censored/Buckets: tolerance = 0
#   Range: tolerance = 1
#
# nominal_codes:
#   Nominal categories that agree only with themselves.
#   Here:
#     6 = Missing
#     7 = Non-Applicable

make_disagreement_function <- function(
    ordinal_positions,
    tolerance = 0,
    nominal_codes = NOMINAL_CODES
) {

  # Ensure category names are available for lookup.
  if (is.null(names(ordinal_positions))) {
    stop("ordinal_positions must be a named numeric vector.")
  }

  ordinal_codes <- as.numeric(names(ordinal_positions))

  # Maximum effective ordinal distance after applying the tolerance.
  #
  # Total:
  #   positions 1-5, tolerance 0
  #   maximum effective distance = 4
  #
  # Range:
  #   positions 1-5, tolerance 1
  #   maximum effective distance = 3
  #
  # Buckets:
  #   positions 1-3, tolerance 0
  #   maximum effective distance = 2
  maximum_effective_distance <-
    max(ordinal_positions) -
    min(ordinal_positions) -
    tolerance

  if (maximum_effective_distance <= 0) {
    stop("The ordinal scale must contain more variation than the tolerance.")
  }

  function(x, y) {

    # The package may pass genuine R missing values to the distance
    # function. These are excluded from the disagreement calculation.
    if (is.na(x) || is.na(y)) {
      return(0)
    }

    # Exact matches always have zero disagreement.
    #
    # This includes:
    #   6 versus 6
    #   7 versus 7
    if (x == y) {
      return(0)
    }

    # Any non-identical comparison involving a nominal category
    # receives maximum disagreement.
    #
    # Examples:
    #   5 versus 6 = 1
    #   6 versus 7 = 1
    #   2 versus 7 = 1
    if (x %in% nominal_codes || y %in% nominal_codes) {
      return(1)
    }

    # Confirm that both remaining values are ordinal categories.
    if (!(x %in% ordinal_codes && y %in% ordinal_codes)) {
      stop(
        paste(
          "Unexpected category supplied to disagreement function:",
          x, "and", y
        )
      )
    }

    # Look up the positions of the two ordinal categories.
    position_x <- ordinal_positions[as.character(x)]
    position_y <- ordinal_positions[as.character(y)]

    raw_distance <- abs(position_x - position_y)

    # Remove the number of steps permitted as complete agreement.
    effective_distance <- max(raw_distance - tolerance, 0)

    # Squared standardized distance gives less weight to small differences
    # and progressively more weight to larger differences
    # The most distant ordinal categories receive disagreement = 1.
    (effective_distance / maximum_effective_distance)^2
  }
}
```


<!-- Primary Outcome (Global IRR) Measure 2/2 -->
<!-- Chance-Corrected Agreement in Four Definitions -->
```{r four-definitions}
# [1] Total agreement:
# Scores 1-5 retain their original ordinal positions.
#
# Examples:
#   1 versus 1 = 0
#   1 versus 2 = 1/16
#   1 versus 3 = 4/16
#   1 versus 5 = 1

total_disagreement <- make_disagreement_function(
  ordinal_positions = c(
    "1" = 1,
    "2" = 2,
    "3" = 3,
    "4" = 4,
    "5" = 5
  ),
  tolerance = 0
)


# [2] Range agreement:
# A difference of one ordinal step is treated as complete agreement.
#
# The tolerance applies only to real scores 1-5. Because codes 6 and 7
# are handled as special nominal categories before this calculation:
#
#   4 versus 5 = 0
#   5 versus 6 = 1
#   6 versus 7 = 1
#
# The remaining ordinal disagreement is rescaled so that
# 1 versus 5 remains the maximum disagreement of 1.

range_disagreement <- make_disagreement_function(
  ordinal_positions = c(
    "1" = 1,
    "2" = 2,
    "3" = 3,
    "4" = 4,
    "5" = 5
  ),
  tolerance = 1
)


# [3] Censored agreement:
# Uses the same ordinal disagreement function as Total.
censored_disagreement <- total_disagreement


# [4] Bucketed agreement:
#
# Original scores are first transformed into:
#   1 and 2 -> Low bucket    (code 1)
#   3       -> Middle bucket (code 2)
#   4 and 5 -> High bucket   (code 3)
#   6       -> Missing       (code 6)
#   7       -> Non-Applicable(code 7)
#
# The three numerical buckets are treated as ordinal positions:
#   Low versus High    = 1.00 disagreement
#   Low versus Middle  = 0.25 disagreement
#   Middle versus High = 0.25 disagreement

bucket_disagreement <- make_disagreement_function(
  ordinal_positions = c(
    "1" = 1,
    "2" = 2,
    "3" = 3
  ),
  tolerance = 0
)
```


<!-- Transformations and recodes specific to the four definitions -->
```{r dataset-transformations}
# Total and Range use the original complete seven-code dataset.
alpha_data_total <- ratings_units_by_raters
alpha_data_range <- ratings_units_by_raters


# Censored removes an entire observation when either rater assigned
# Missing (6) or Non-Applicable (7).
alpha_data_censored <- ratings_units_by_raters[
  apply(
    ratings_units_by_raters,
    MARGIN = 1,
    FUN = function(observation) {
      !any(observation %in% NOMINAL_CODES)
    }
  ),
  ,
  drop = FALSE
]


# Recode the original categories for the bucketed analysis.
recode_to_bucket <- function(x) {
  case_when(
    x %in% c(1, 2) ~ 1,
    x == 3          ~ 2,
    x %in% c(4, 5) ~ 3,
    x == 6          ~ 6,
    x == 7          ~ 7,
    is.na(x)        ~ NA_real_,
    TRUE            ~ NA_real_
  )
}

alpha_data_bucket <- matrix(
  recode_to_bucket(as.vector(ratings_units_by_raters)),
  nrow = nrow(ratings_units_by_raters),
  ncol = ncol(ratings_units_by_raters),
  dimnames = dimnames(ratings_units_by_raters)
)


# Confirm that each analysis dataset contains only permitted codes, has no
# unexpected missing values, and retains eligible observations from raters
validate_alpha_data <- function(
    data,
    valid_codes,
    analysis_name,
    allow_na = FALSE
) {

  observed_codes <- sort(unique(as.vector(data)))
  observed_nonmissing <- observed_codes[!is.na(observed_codes)]

  invalid_codes <- setdiff(observed_nonmissing, valid_codes)

  if (length(invalid_codes) > 0) {
    stop(
      analysis_name,
      " contains invalid codes: ",
      paste(invalid_codes, collapse = ", ")
    )
  }

  if (!allow_na && anyNA(data)) {
    stop(analysis_name, " contains unexpected genuine NA values.")
  }

  if (ncol(data) < 2) {
    stop(analysis_name, " requires at least two raters.")
  }

  if (nrow(data) == 0) {
    stop(analysis_name, " contains no eligible observations.")
  }

  invisible(TRUE)
}


# Validate each dataset against its definition-specific codes
validate_alpha_data(
  alpha_data_total,
  valid_codes = 1:7,
  analysis_name = "Total dataset"
)

validate_alpha_data(
  alpha_data_range,
  valid_codes = 1:7,
  analysis_name = "Range dataset"
)

validate_alpha_data(
  alpha_data_censored,
  valid_codes = 1:5,
  analysis_name = "Censored dataset"
)

validate_alpha_data(
  alpha_data_bucket,
  valid_codes = c(1:3, 6:7),
  analysis_name = "Bucket dataset"
)
```


```{r wrapper-function}
# Run Krippendorff's Alpha calculation using the supplied disagreement
# function and return a fitted model with information about the dataset
calculate_alpha <- function(
    data,
    disagreement_function,
    analysis_name,
    confint = TRUE
) {

  # Use analytical estimation with reproducible, nonparallel execution
  fit <- krippendorffs.alpha(
    data = data,
    level = disagreement_function,
    method = "analytical",
    confint = confint,
    verbose = FALSE,
    control = list(parallel = FALSE)
  )

  list(
    name = analysis_name,
    observations = nrow(data),
    raters = ncol(data),
    fit = fit
  )
}
```


```{r alpha-calls}
# Estimate Krippendorff's alpha once for each agreement definition
alpha_total <- calculate_alpha(
  data = alpha_data_total,
  disagreement_function = total_disagreement,
  analysis_name = "Total"
)

alpha_range <- calculate_alpha(
  alpha_data_range,
  range_disagreement,
  "Range"
)

alpha_censored <- calculate_alpha(
  alpha_data_censored,
  censored_disagreement,
  "Censored"
)

alpha_bucket <- calculate_alpha(
  alpha_data_bucket,
  bucket_disagreement,
  "Buckets"
)

# Tabulate the estimates with 95% CI for reporting
alpha_summary <- data.frame(
  Definition = c("Total", "Range", "Censored", "Bucketed"),
  Observations = c(
    alpha_total$observations,
    alpha_range$observations,
    alpha_censored$observations,
    alpha_bucket$observations
  ),
  Alpha = c(
    alpha_total$fit$alpha.hat,
    alpha_range$fit$alpha.hat,
    alpha_censored$fit$alpha.hat,
    alpha_bucket$fit$alpha.hat
  ),
  Lower_95_CI = c(
    alpha_total$fit$L,
    alpha_range$fit$L,
    alpha_censored$fit$L,
    alpha_bucket$fit$L
  ),
  Upper_95_CI = c(
    alpha_total$fit$U,
    alpha_range$fit$U,
    alpha_censored$fit$U,
    alpha_bucket$fit$U
  )
)

# Display
kable(
  alpha_summary,
  digits = 4,
  caption = "Summary of Krippendorff's alpha under each agreement definition."
)
```



# Secondary Analyses

## Crude Agreement by Domain
<!-- Secondary Outcome 1/2 - Raw Agreement by Domain -->
<!-- Defined as it is for primary outcome, only applied to each individual
domain (e.g., "Eligiblity", "Recruitment") -->
```{r}
# Apply the same four crude-agreement definitions within each domain.
domain_agreement_one_definition <- function(data, definition) {
  # remove observations with a nominal code for censored agreement
  if (definition == "Censored") {
    data <- data |>
      filter(
        !observer_1 %in% NOMINAL_CODES,
        !observer_2 %in% NOMINAL_CODES
      )
  }
  
  # count agreements and eligible observations seperately within each domain
  data |>
    group_by(domain_number, domain) |>
    summarise(
      Agreements = case_when(
        definition == "Total" ~ sum(observer_1 == observer_2),
        # one-point differences are full agreement for range agreement
        definition == "Range" ~ sum(
          if_else(
            observer_1 %in% ORDINAL_CODES & observer_2 %in% ORDINAL_CODES,
            abs(observer_1 - observer_2) <= 1,
            observer_1 == observer_2
          )
        ),
        definition == "Censored" ~ sum(observer_1 == observer_2),
        # for bucketed agreement, both scores must be in the same bucket
        # to fully agree
        definition == "Bucketed" ~ sum(
          bucket_scores(observer_1) == bucket_scores(observer_2)
        )
      ),
      Total = n(),
      .groups = "drop"
    ) |>
    mutate(
      Definition = definition,
      Crude_Agreement = Agreements / Total,
      .after = domain
    )
}

# Apply the function to all four definitions and arrange results
domain_agreement_summary <- map_dfr(
  c("Total", "Range", "Censored", "Bucketed"),
  ~ domain_agreement_one_definition(ratings_tidy, .x)
) |>
  arrange(domain_number, factor(
    Definition,
    levels = c("Total", "Range", "Censored", "Bucketed")
  )) |>
  select(
    domain_number,
    domain,
    Definition,
    Agreements,
    Total,
    Crude_Agreement
  )

# (10 domains x 4 agreement definitions) = 40 results
if (nrow(domain_agreement_summary) != 40L) {
  stop("Expected 40 domain-level crude-agreement results.")
}

# compare to preflight results
expected_domain_agreement <- read.csv(
  repo_path("outputs", "preflight", "domain_results_preflight.csv"),
  stringsAsFactors = FALSE
) |>
  transmute(
    domain_number,
    domain,
    Definition = definition,
    Agreements = agreements,
    Total = observations,
    Crude_Agreement = crude_agreement_percent / 100
  ) |>
  arrange(domain_number, factor(
    Definition,
    levels = c("Total", "Range", "Censored", "Bucketed")
  ))

domain_comparison <- all.equal(
  domain_agreement_summary,
  expected_domain_agreement,
  tolerance = 0.0000001,
  check.attributes = FALSE
)
# stop if the results don't match
if (!isTRUE(domain_comparison)) {
  stop(
    "Domain-level agreement does not match the verified preflight table: ",
    paste(domain_comparison, collapse = "; ")
  )
}

# display results
kable(
  domain_agreement_summary |>
    select(-domain_number),
  digits = 4,
  caption = "Crude agreement by PRECIS-2 domain and agreement definition.",
  longtable = TRUE,
  booktabs = TRUE
)
```


## Mean Signed Deviation from Consensus
<!-- Secondary Outcome 2/2 - MSD from Consensus -->
<!-- Calculate the average directional difference from consensus score
to observe asymmetry and/or bias in the scoring tendencies of reviewers -->
```{r}
# Calculate each reviewer's signed and absolute deviation from consensus
# Restrict the analysis to observations where both reviewers and consensus
# have ordinal scores (1-5), excluding codes 6 and 7
ratings_with_consensus <- ratings_tidy |>
  select(
    observation_id, trial, domain_number, domain,
    observer_1, observer_2
  ) |>
  left_join(
    consensus_tidy |>
      select(observation_id, consensus),
    by = "observation_id"
  )

# retain a common set of eligible observations for both reviewers
eligible_consensus_ratings <- ratings_with_consensus |>
  filter(
    observer_1 %in% ORDINAL_CODES,
    observer_2 %in% ORDINAL_CODES,
    consensus %in% ORDINAL_CODES
  )

# merge into one table
# calculate and summarize magnitude and direction of deviation
reviewer_deviation_summary <- bind_rows(
  eligible_consensus_ratings |>
    transmute(
      Reviewer = "Aarian",
      reviewer_score = observer_1,
      consensus
    ),
  eligible_consensus_ratings |>
    transmute(
      Reviewer = "Merrick",
      reviewer_score = observer_2,
      consensus
    )
) |>
  mutate(deviation = reviewer_score - consensus) |>
  group_by(Reviewer) |>
  summarise(
    Observations = n(),
    Signed_Deviation_Sum = sum(deviation),
    Absolute_Deviation_Sum = sum(abs(deviation)),
    Mean_Signed_Deviation = mean(deviation),
    Mean_Absolute_Deviation = mean(abs(deviation)),
    .groups = "drop"
  )

# record independently verified totals used for regression testing
expected_deviation_totals <- tibble(
  Reviewer = c("Aarian", "Merrick"),
  Observations = c(211L, 211L),
  Signed_Deviation_Sum = c(-3, 6),
  Absolute_Deviation_Sum = c(97, 92)
)

# select comparable values and standardize reviewer order before validation
observed_deviation_totals <- reviewer_deviation_summary |>
  select(
    Reviewer,
    Observations,
    Signed_Deviation_Sum,
    Absolute_Deviation_Sum
  ) |>
  arrange(Reviewer)

expected_deviation_totals <- expected_deviation_totals |>
  arrange(Reviewer)

# stop the program if any unexplained changes to deviation occur
deviation_comparison <- all.equal(
  observed_deviation_totals,
  expected_deviation_totals,
  check.attributes = FALSE
)

if (!isTRUE(deviation_comparison)) {
  stop(
    "Reviewer-deviation totals do not match the finalized Step 1 results: ",
    paste(deviation_comparison, collapse = "; ")
  )
}

# display reviewer-specific deviations in a table
kable(
  reviewer_deviation_summary |>
    transmute(
      Reviewer,
      `Eligible observations` = Observations,
      `Mean signed deviation` = Mean_Signed_Deviation,
      `Mean absolute deviation` = Mean_Absolute_Deviation
    ),
  digits = 6,
  caption = "Reviewer deviation from consensus after excluding codes 6 and 7.",
  booktabs = TRUE
)
```


# Reproducibility Information
<!-- Step 1 - Produce Reusable CSV Files -->
```{r export-results, include=FALSE}
# Create the results directory if it doesn't already exist
results_directory <- repo_path("outputs", "results")
dir.create(results_directory, recursive = TRUE, showWarnings = FALSE)

# Export the principal and secondary analysis results as reusable CSV files
write.csv(
  agreement_summary,
  file.path(results_directory, "global-crude-agreement.csv"),
  row.names = FALSE
)
write.csv(
  domain_agreement_summary,
  file.path(results_directory, "domain-crude-agreement.csv"),
  row.names = FALSE
)
write.csv(
  reviewer_deviation_summary,
  file.path(results_directory, "reviewer-deviation-from-consensus.csv"),
  row.names = FALSE
)
write.csv(
  alpha_summary,
  file.path(results_directory, "krippendorff-alpha.csv"),
  row.names = FALSE
)

# Record information about the R session and package environment
writeLines(
  capture.output(sessionInfo()),
  file.path(results_directory, "session-info.txt")
)
```

<!-- Step 2 - Documenting the Environment -->
```{r}
# Summarize the principal software and package versions used for this run
reproducibility_summary <- tibble(
  Component = c(
    "R",
    "Platform",
    "krippendorffsalpha",
    "dplyr",
    "rmarkdown",
    "knitr"
  ),
  Version = c(
    paste(R.version$major, R.version$minor, sep = "."),
    R.version$platform,
    as.character(packageVersion("krippendorffsalpha")),
    as.character(packageVersion("dplyr")),
    as.character(packageVersion("rmarkdown")),
    as.character(packageVersion("knitr"))
  )
)

# Display the software environment used
kable(
  reproducibility_summary,
  caption = paste(
    "Software environment. Full session information is written to",
    "outputs/results/session-info.txt."
  ),
  booktabs = TRUE
)
```
